---
title: "ps5-markdown-plots"
author: "chloe abrahams"
date: "2023-02-17"
output: html_document
---
## 1: Load and Check Data
```{r}
library(tidyverse)
```

2: 
```{r}
gapminder <- read_delim("C:/Users/chloe/OneDrive/UW/INFO201/data/gapminder.csv.bz2")
gapminder %>% 
  dim()
```
This dataset contains `r nrow(gapminder)` rows and
  `r ncol(gapminder)` columns.

3: 
```{r}
gapminder %>% 
  sample_n(6)
```
the data looks good 


## 2: Descriptive Statistics
1:
```{r}
num_iso3 <- gapminder %>% 
  distinct(iso3) %>% 
  nrow()

num_iso2 <- gapminder %>% 
  distinct(iso2) %>% 
  nrow()

num_names <- gapminder %>% 
  distinct(name) %>% 
  nrow()

cat("There are", num_iso3, "unique iso3 codes in the dataset.\n")
cat("There are", num_iso2, "unique iso2 codes in the dataset.\n")
cat("There are", num_names, "unique country names in the dataset.\n")
```
2a: 
```{r}
iso2_names <- gapminder %>% 
  group_by(iso2) %>% 
  summarize(num_names = n_distinct(name))

multi_names <- iso2_names %>% 
  filter(num_names > 1)

multi_names_countries <- gapminder %>% 
  filter(!is.na(iso2) %in% !is.na(multi_names$iso2)) %>% 
  select(iso2, name) %>% 
  distinct() %>% 
  arrange(iso2, name)

print(multi_names_countries)
```

2b: 
```{r}
name_iso3 <- gapminder %>% 
  group_by(name) %>% 
  summarize(num_iso3 = n_distinct(iso3))

multi_iso3_names <- name_iso3 %>% 
  filter(num_iso3 > 1)

cat("\nCountry names with more than one iso3 code:\n")
multi_iso3_countries <- gapminder %>% 
  filter(!is.na(name) %in% !is.na(multi_iso3_names$name)) %>% 
  select(name, iso3) %>% 
  distinct() %>% 
  arrange(name, iso3)
print(multi_iso3_countries)
```

3: 
```{r}
max_year <- max(gapminder$time, na.rm = TRUE)
cat("The max year in the dataset:", max_year, "\n")

min_year <- min(gapminder$time, na.rm = TRUE)
cat("The min year in the dataset:", min_year, "\n")
```
## CO2 Emissions
1: 
```{r}
gapminder %>% 
  group_by(time) %>% 
  summarize(num_missing_co2 = sum(is.na(co2)), num_missing_co2_PC = sum(is.na(co2_PC))) %>% 
  mutate(total_missing = num_missing_co2 + num_missing_co2_PC) %>% 
  arrange(desc(total_missing)) %>% 
  head(5)
```



2: 
```{r}
selected_countries <- c("United States of America", "China", "India", "Japan", "Germany")
gapminder %>% 
  filter(!is.na(co2)) %>% 
  filter(name %in% selected_countries) %>% 
  ggplot(aes(time, co2, col = factor(name))) + 
  geom_line() + 
  labs(x = "Year", y = "Total CO2 Emissions (million metric tons)", 
       col = "Country")
```



3: 
```{r}
selected_countries <- c("United States of America", "China", "India", "Japan", "Germany")
gapminder %>% 
  filter(!is.na(co2_PC)) %>% 
  filter(name %in% selected_countries) %>% 
  ggplot(aes(time, co2_PC, col = factor(name))) + 
  geom_line() + 
  labs(x = "Year", y = "Total CO2 Emissions per Capita (million metric tons)", 
       col = "Country")
```



All countries save for China and India increased their CO2 emissions 
when analyzing emissions per capita. Because per capita means per 
person, this speaks to the size of China and India's populations - 
the other countries are releasing more CO2 emissions per person. 

4. 
```{r}
continent_avg <- gapminder %>% 
  filter(!is.na(co2_PC), !is.na(region)) %>% 
  group_by(region) %>% 
  summarize(avg_co2_PC = mean(co2_PC, na.rm = TRUE))

print(continent_avg)
```



Europe has the highest per capita emissions out of all the regions, and the 
more developed continents (Europe, Americas, Asia) have more average 
emissions than Oceania and Africa, with less development and/or less people. 


5. 
```{r}
data_filtered <- gapminder %>% 
  filter(!is.na(co2), !is.na(region)) %>% 
  filter(time %in% c(1960, 2016))

continent_avg <- data_filtered %>% 
  group_by(region, time) %>% 
  summarize(avg_co2_PC = mean(co2_PC, na.rm = TRUE))

ggplot(continent_avg, aes(x = region, y = avg_co2_PC, fill = as.factor(time))) +
  geom_col(position = "dodge") +
  labs(x = "Continent", y = "Average CO2 Emissions per Capita", fill = "Year") +
  ggtitle("Average CO2 Emissions per Capita Across Continents") +
  theme_minimal() 
```



6. 
Instead of using 2019, I chose to use 2016 because there 
were less NA values. 
```{r}
data_2016 <- gapminder %>% 
  filter(time == 2016)

grouped_data <- data_2016 %>% 
  group_by(region, name) %>% 
  summarize(avg_co2_PC = mean(co2_PC, na.rm = TRUE))

ordered_data <- grouped_data %>% 
  arrange(region, desc(avg_co2_PC))

top_three <- ordered_data %>% 
  group_by(region) %>% 
  top_n(3, wt = avg_co2_PC)

bottom_three <- ordered_data %>% 
  group_by(region) %>% 
  top_n(-3, wt = avg_co2_PC)

print(top_three)
print(bottom_three)

```
## GDP per Capita

1: 
```{r}
data_1960 <- gapminder %>%
  filter(!is.na(region), !is.na(GDP_PC), !is.na(lifeExpectancy)) %>% 
  filter(time == 1960)

ggplot(data = data_1960, aes(x = GDP_PC, y = lifeExpectancy, color = region, 
                             size = 1)) +
  geom_point() +
  scale_color_manual(values = c("#FF6666", "#66CCFF", "#66FF66", "#FFCC66", "#CC66FF", "#CCCCCC")) +
  labs(x = "GDP per capita", y = "Life expectancy", title = "GDP per capita vs. Life expectancy by country (1960)") +
  theme_bw()

```


The countries with a lower GDP per capita cluster together with the lowest 
life expectancies, while the countries with the highest GDP per capitas are
more spread out iwth the highest life expectancies. 

2. 
```{r}
data_2019 <- gapminder %>%
  filter(!is.na(region), !is.na(GDP_PC), !is.na(lifeExpectancy)) %>% 
  filter(time == 2019)

ggplot(data = data_2019, aes(x = GDP_PC, y = lifeExpectancy, color = region, 
                             size = 1)) +
  geom_point() +
  scale_color_manual(values = c("#FF6666", "#66CCFF", "#66FF66", "#FFCC66", "#CC66FF", "#CCCCCC")) +
  labs(x = "GDP per capita", y = "Life expectancy", title = "GDP per capita vs. Life expectancy by country (2019)") +
  theme_bw()
```



3. There are significantly less countries clustered in the lowest range of 
life expectancy, 0-60, with the majority now clustering between 65-80. 
This shows a global improvement in health with an increase in overall 
global life expectancy, indicating major global development.

4.
```{r}
data_1960 <- gapminder %>% 
  filter(time == 1960)

data_2019 <- gapminder %>% 
  filter(time == 2019)

life_exp_1960 <- data_1960 %>% 
  group_by(region) %>% 
  summarize(avg_life_exp = mean(lifeExpectancy, na.rm = TRUE))

life_exp_2019 <- data_2019 %>% 
  group_by(region) %>% 
  summarize(avg_life_exp = mean(lifeExpectancy, na.rm = TRUE))

life_exp_1960
life_exp_2019
```
Yes, there is a significant increase between the life expectancies of 
1960 and 2019, consistent with what is shown on the 2 scatter plots. 

5. 
```{r}
grouped_data <- gapminder %>% 
  filter(time == 1960 | time == 2019, !is.na(time), !is.na(region)) %>% 
  group_by(region, time) %>% 
  summarize(avg_le = mean(lifeExpectancy, na.rm = TRUE))

grouped_data_lag <- grouped_data %>% 
  group_by(region) %>% 
  mutate(lag_le = lag(avg_le, order_by = time))

growth_data <- grouped_data_lag %>% 
  filter(!is.na(lag_le)) %>%
  mutate(growth_rate = (avg_le - lag_le) / lag_le)

avg_growth <- growth_data %>% 
  group_by(region) %>% 
  summarize(avg_growth_rate = mean(growth_rate, na.rm = TRUE)) %>% 
  arrange(desc(avg_growth_rate))

print(avg_growth)
```
Africa and Asia have the highest growth rates, possible due to the fact
that they include some of the least developed regions in the world, where
high fertility rates are common. On the other hand, the overall more developed
Americas and Europe have the lowest growth rate, likely due to low fertility
rates. 

6. 
```{r}
data_1960 <- gapminder %>% 
  filter(time == 1960, !is.na(GDP_PC))

data_2019 <- gapminder %>% 
  filter(time == 2019, !is.na(GDP_PC))

combined_data <- rbind(data_1960, data_2019)
combined_data$time <- factor(combined_data$time)
ggplot(combined_data, aes(x = GDP_PC, fill = time)) +
  geom_histogram(binwidth = 5000, alpha = 0.5, position = "dodge") +
  scale_fill_manual(values = c("red3", "lightblue")) +
  xlab("GDP per capita") +
  ylab("Count") +
  ggtitle("Histogram of GDP per capita for 1960 and 2019")
```



7. 
```{r}
ranked_data <- gapminder %>% 
  group_by(time) %>% 
  mutate(rank_LE = rank(desc(lifeExpectancy), na.last = "keep", ties.method ="min")) %>% 
 filter(time %in% c(1960, 2019))

us_data <- ranked_data %>% 
  filter(name == "United States of America")

min_ranks <- us_data %>% 
  summarize(min_rank = min(rank_LE))

print(min_ranks)

```
my code is running correctly, i think there might be an issue with 
the data? i couldn't figure out why I wasn't getting 17

8. 
```{r}
data_sub <- gapminder %>% 
  filter(time %in% c(1960, 2019)) 

n_countries <- data_sub %>% 
  group_by(time) %>% 
  summarize(n_countries = n_distinct(name))

#print(n_countries)

ranked_data <- data_sub %>% 
  group_by(time) %>% 
 # mutate(rank_LE = rank(desc(lifeExpectancy), na.last = "keep"))
  mutate(rank_LE = rank(desc(lifeExpectancy), na.last = TRUE))

rel_rank <- ranked_data %>% 
  inner_join(n_countries, by = "time") %>% 
  mutate(rel_rank = rank_LE / n_countries)

us_data <- rel_rank %>% 
  filter(name == "United States of America") %>% 
  select(name, rel_rank)
#summarise(us_data = min(rel_rank))

print(us_data)

```
this corresponds with my answer to #7, for the same reason, 
I'm just slightly off 
